String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-15 2023-1-19/sections/cefa.tex
typo
typo removement

%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, I, F)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, I, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers.
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha \in \Phi(R)$ is an LIA formula specifying an accepting condition.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}
For a CEFA $\aut$, we use $R_\aut$ to denote the set of registers of $\aut$. 
%The definition of CEFA above is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
%are attached to final states.

@@

The \emph{union} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \cup \aut_2$, is defined in a slightly more involved manner than NFA, as a result of the technical choice that the accepting conditions of CEFA do not distinguish final states. Two fresh auxiliary registers, say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ or from $\aut_2$. Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$, where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
\item $\delta'$ is the union of $\{(q'_0, a, q'_1, (\myvec{v_1}, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, $\{(q_1, a, q'_1, (\myvec{v_1}, \myvec{0}, 0, 0)) \mid (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, and $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 0)) \mid (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, where $\myvec{0}$ in $(\myvec{v_1}, \myvec{0}, 1,0)$ is a vector of updates on $R_2$ and $1,0$, the last two components of the vector, denote the updates on $r'_1$ and $r'_2$, similarly for $(\myvec{0}, \myvec{v_2}, 0, 1)$, and so on.

\item $F'$ and $\alpha'$ are defined as follows, 
\begin{itemize}
\item if $I_1 \cap F_1 \neq \emptyset$ or $I_2 \cap F_2 \neq \emptyset$, then $F' = F_1 \cup F_2 \cup \{q'_0\}$ and $\alpha' = (r'_1 = 0 \wedge r'_2 = 0) \vee (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$, 
%
\item otherwise, $F'=F_1 \cup F_2$ and $\alpha' = (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$.
\end{itemize}
\end{itemize}
From the construction, we know that 
$$\Lang(\aut_1 \cup \aut_2) = \{(w; \myvec{v}) \mid \myvec{v} \in \Int^{R'}, (w; \prj_{R_1}(\myvec{v})) \in \Lang(\aut_1) \mbox{ or } (w; \prj_{R_2}(\myvec{v})) \in \Lang(\aut_2)\}.$$

>>>>

%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, I, F, \alpha)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, I, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers.
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha \in \Phi(R)$ is an LIA formula specifying an accepting condition.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}
For a CEFA $\aut$, we use $R_\aut$ to denote the set of registers of $\aut$. 
%The definition of CEFA above is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
%are attached to final states.

@@

The \emph{union} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \cup \aut_2$, is defined in a slightly more involved manner than NFA, as a result of the technical choice that the accepting conditions of CEFA do not distinguish final states. Two fresh auxiliary registers, say $r'_1,r'_2 \not \in R_1 \cup R_2$, are introduced so that the accepting condition knows whether a run is from $\aut_1$ or from $\aut_2$. Specifically, $\aut_1 \cup \aut_2 = (R', Q', \Sigma, \delta', I', F', \alpha')$, where 
\begin{itemize}
\item $R' = R_1 \cup R_2 \cup \{r'_1, r'_2\}$, $Q' = Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $I' = \{q'_0\}$, 
%
\item $\delta'$ is the union of $\{(q'_0, a, q'_1, (\myvec{v_1}, \myvec{0}, 1,0)) \mid \exists q_1 \in I_1.\ (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, $\{(q'_0, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 1)) \mid \exists q_2 \in I_2.\ (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, $\{(q_1, a, q'_1, (\myvec{v_1}, \myvec{0}, 0, 0)) \mid (q_1, a, q'_1, \myvec{v_1}) \in \delta_1\}$, and $\{(q_2, a, q'_2, (\myvec{0}, \myvec{v_2}, 0, 0)) \mid (q_2, a, q'_2, \myvec{v_2}) \in \delta_2\}$, where $(\myvec{v_1}, \myvec{0}, 1,0)$ is a vector that updates $R_1$ by $\myvec{v_1}$, updates $R_2$ by $\myvec{0}$, and updates $r'_1,r'_2$ by $1,0$ respectively. Similarly for $(\myvec{0}, \myvec{v_2}, 0, 1)$, and so on.

\item $F'$ and $\alpha'$ are defined as follows, 
\begin{itemize}
\item if either $I_1 \cap F_1 \neq \emptyset$ and $\alpha_1[\myvec{0}/R_1]$ is true, or $I_2 \cap F_2 \neq \emptyset$ and $\alpha_2[\myvec{0}/R_2]$ is true, then $F' = F_1 \cup F_2 \cup \{q'_0\}$ and $\alpha' = (r'_1 = 0 \wedge r'_2 = 0) \vee (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$, 
%
%\item otherwise, if $I_2 \cap F_2 \neq \emptyset$, then $F' = F_1 \cup F_2 \cup \{q'_0\}$ and $\alpha' = (r'_1 = 0 \wedge r'_2 = 0 \wedge \alpha_2) \vee (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$, 
%
\item otherwise, $F'=F_1 \cup F_2$ and $\alpha' = (r'_1=1 \wedge \alpha_1) \vee (r'_2=1 \wedge \alpha_2)$.
\end{itemize}
\end{itemize}
From the construction, we know that 
$$
\Lang(\aut_1 \cup \aut_2) = 
\left\{(w; \myvec{v}) \left \vert 
\begin{array}{l}
%\myvec{v} \in \Int^{R'}, 
(w; \prj_{R_1}(\myvec{v})) \in \Lang(\aut_1) \mbox{ and } \prj_{R_2}(\myvec{v}) = \myvec{0}, \mbox{ or } \\
(w; \prj_{R_2}(\myvec{v})) \in \Lang(\aut_2) \mbox{ and } \prj_{R_1}(\myvec{v}) = \myvec{0}
\end{array}
\right.
\right\}.
$$

### 2023-05-15 2023-1-19/sections/overview.tex
typo removement

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut_1 \cap \aut_2 \cap \aut_3$, producing a cost $(v_1, v_2, v_3, v_4)$ that satisfies $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $v_4 > 120$. 
It is not hard to observe that the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is independent of the characters of $\aut_1 \cap \aut_2 \cap \aut_3$.  Therefore, the characters in $\aut_1 \cap \aut_2 \cap \aut_3$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$ (see Figure~\ref{fig:overview:product:reduced}(a)), where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$. Then the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e. numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $\myvec{v'} = \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$. Intuitively, $\myvec{v'}$ here is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight for $\myvec{v}$ is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

%
Let $\costset = \{\myvec{v_1}, \cdots, \myvec{v_m}\}$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi_\cB(\anivar_1, \cdots, \anivar_m)$ can be computed to define the Parikh image of strings that are accepted by $\cB$, where $\anivar_1, \cdots, \anivar_m$ are the integer variables to denote the number of occurrences of $\myvec{v_1}, \cdots, \myvec{v_m}$.
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\begin{array}{l}
\psi_\cB(\anivar_1, \cdots, \anivar_m) \wedge \bigwedge \limits_{1 \le j \le 4} r_j = \sum \limits_{1\le k \le m}  \anivar_k v_{k, j}\ \wedge \\
1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_4 > 120.
\end{array}
\end{equation}
%where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. 
Since the satisfiability of LIA formulas can be solved by the off-the-shelf SMT solvers for LIA, the satisfiability of the string constraint in~(\ref{eqn-running}) can thus be solved.

Nevertheless, it turns out that the algorithm illustrated above is insufficient to solve the string constraints with regex-counting and string-length efficiently, since the sizes of NFA obtained from CEFA can still be big, resulting  in  the LIA formulas of big sizes, while the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}). 
To tackle this issue, we propose techniques to reduce the sizes of NFA over the alphabet $\costset$.

@@

We determinize $\cB$, apply the minimization algorithm to the resulting deterministic finite automaton (DFA), and obtain a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b). 
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
Although the determinization and minimization may not reduce the sizes of NFA in the worst case, the experiment results show us that they do reduce the sizes of NFA constructed from the regular expressions in practice. 
%(See Table~\ref{tab:results_heuristics}, where adding the size-reduction techniques reduces the running time for 46.4\%.) 
%thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
%Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
%Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
%\begin{equation}\label{eqn-LIA-reduced}
%\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
%\end{equation}
%where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

>>>>

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut_1 \cap \aut_2 \cap \aut_3$, producing a cost $(v_1, v_2, v_3, v_4)$ that satisfies $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $v_4 > 120$. 
It is not hard to observe that the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is independent of the characters of $\aut_1 \cap \aut_2 \cap \aut_3$.  Therefore, the characters in $\aut_1 \cap \aut_2 \cap \aut_3$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$ (see Figure~\ref{fig:overview:product:reduced}(a)), where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$. Then the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e. numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $\myvec{v'} = (v'_1, v'_2, v'_3, v'_4) =  \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$. Intuitively, $\myvec{v'}$ here is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight for $\myvec{v}$ is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

%
Let $\costset = \{\myvec{v_1}, \cdots, \myvec{v_m}\}$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi_\cB(\anivar_1, \cdots, \anivar_m)$ can be computed to define the Parikh image of strings that are accepted by $\cB$, where $\anivar_1, \cdots, \anivar_m$ are the integer variables to denote the number of occurrences of $\myvec{v_1}, \cdots, \myvec{v_m}$.
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\begin{array}{l}
\psi_\cB(\anivar_1, \cdots, \anivar_m) \wedge \bigwedge \limits_{1 \le j \le 4} r_j = \sum \limits_{1\le k \le m}  \anivar_k v_{k, j}\ \wedge \\
1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_4 > 120.
\end{array}
\end{equation}
%where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. 
Finally, the satisfiability of the LIA formula in~(\ref{eqn-LIA}) can be solved by the off-the-shelf SMT solvers for LIA.
%Since the satisfiability of LIA formulas can be solved by the off-the-shelf SMT solvers for LIA, the satisfiability of the string constraint in~(\ref{eqn-running}) can thus be solved.

Nevertheless, the algorithm illustrated above is insufficient to solve the string constraints with regex-counting and string-length efficiently, since the sizes of NFA obtained from CEFA can still be big, resulting  in  the LIA formulas of big sizes, while the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}). 
To tackle this issue, we propose techniques to reduce the sizes of NFA over the alphabet $\costset$.

@@

Specifically, to reduce the sizes of the NFA, we determinize $\cB$, apply the minimization algorithm to the resulting deterministic finite automaton (DFA), and obtain a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b). 
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
Although the determinization and minimization may not reduce the sizes of NFA in the worst case, the experiment results show us that they do reduce the sizes of NFA constructed from the regular expressions in practice. 
%(See Table~\ref{tab:results_heuristics}, where adding the size-reduction techniques reduces the running time for 46.4\%.) 
%thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
%Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
%Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
%\begin{equation}\label{eqn-LIA-reduced}
%\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
%\end{equation}
%where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.
